$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-04-30 qdpm_paper.tex
Reorganized results sections

\newcommand{\figref}[1]{Fig. \ref{#1}}
\newcommand{\secref}[1]{Sec. \ref{#1}}

@@

\newcommand{\HTC}{H_\mathrm{TC}}

@@

\title{Topological order in the quantum dimer-pentamer model}

@@

Over the last several decades, experimental and theoretical work has demonstrated that strongly interacting quantum many-body systems support exotic quantum phases phases of matter not described by the Landau symmetry breaking paradigm. One class of such quantum phases of matter are topologically ordered quantum liquids~\cite{Wen1990}, first realized experimentally via the fractional quantum Hall effect~\cite{FQHE????}. Despite the absence of conventional symmetry breaking and a local order parameter, topological quantum liquids possess a non-trivial quantum order that distinguishes them from trivial liquid phases~\cite{Nayak2008}. Although not detectable by local operators, topological order arises due to a non-local entanglement structure that leaves a signatures in the bipartite entanglement entropy~\cite{Levin2006a,Kitaev2006b}. Such topological phases provide the basis for a proposal for fault-tolerant quantum information processing~\cite{Freedman2001,Kitaev2003}.

Because of the exotic nature of topological phases and their potential use for quantum information processing, there is great interest in the development of theoretical models which display these phases. Such models offer the possibility of furthering the understanding of these phases as well as providing possibly candidate experimental systems in which to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Martinis2015}. A variety of exactly soluble lattice models are known to support ground states with topological order that ranges from the simplest variety, so-called $Z_2$ topological order~\cite{Kitaev2003,Wen2003}, to those with non-Abelian anyonic excitations~\cite{Levin2005a,Kitaev2006a}. These exactly soluble models generally involve complex many-body interactions, and thus it is desirable to find simpler models supporting topological order that are closer to what is accessible experimentally.

Among the simplest class of models displaying topological quantum liquid phases are those with local geometric constraints; such constraints inhibit the formation of symmetry breaking local order and given sufficient dynamics, such models may have symmetric liquid ground states. The simplest such geometrically constrained model is the quantum dimer model (QDM) describing dimer degrees of freedom living on the links of a lattice~\cite{Rokhsar1988,Moessner2008}; the hard-core, fully-packed QDM has a local constraint that requires exactly one dimer touching each vertex.

@@

While on non-bipartite lattices, the QDM posses a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the QDM is relaxed to a parity constraint such that each vertex much have an even (odd) number of dimers touching each vertex, the simplest non-trivial dynamics lead to the (odd) toric code which has a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in dimer-based models is intimately related to the local constraint which imposes a local gauge symmetry in the low energy Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local (Ising) Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases; the local constraints can often be enforced by a local potential energy penalty and the required dynamics are generally expected to arise at the lowest order non-trivial order in perturbation theory. We will refer to such models with dimer degrees of freedom on links and local constraints at vertices as \emph{generalized quantum dimer modes} (GQDMs). Previous work as demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. Defining $n_v$ as the number of dimers that touch a vertex, we can distinguish between fixed number constraints of the form $n_v = n_0$ and fixed parity constraints of the form $n_v\mod 2 = n_P$ (see \figref{fig:example_local_constraints}).

@@

Because of the local constraints, all states in the Hilbert space of a GQDM are invariant under local gauge transformations a the vertices, $G_v$ of the general form~\cite{Moessner2001}
\begin{equation}
G_v = e^{i \alpha \left( n_v -c_0 \right)}; \notag
\end{equation}
where the $c_0$ and $\alpha$ depend on the constraint.  
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}
If a defect is introduced which violates the local constraint at a vertex, this can be viewed as an ``electric charge'' living on the vertex, within the context of the original GQDM. One may determine if such electric charges are confined by introducing a pair of such defects and computing their spatial correlation function, thus determining the confined or deconfined nature of the GQDM ground state.

The gauge symmetry allows one to construct a gauge invariant string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv -\frac{c_0}{4} \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) 
\end{equation}
and the total charge $Q_\Gamma$ enclosed in a closed loop $\Gamma$ as
\begin{equation}
Q_\Gamma \equiv \Phi_\Gamma+ \tilde{\Phi}_\Gamma.
\end{equation}
(Alternately can think of the source of $\Phi_\Gamma$ to be static charges of charge $+c_0$ ($-c_0$) sitting at the verticies of the $A$($B$) sub-lattice.).

We define the string operator is defined as
\begin{equation}
S_\gamma \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic charges.

@@

The two unique fixed-number constraints on the square lattice $n_0 = 1$ and $n_0=2$ lead to the QDM and the quantum full-packed loop model (QFPLM), respectively. For such GQDMs with a number constraint, we write the gauge transformation as:
\begin{equation}
G_v = e^{i \alpha \left( n_v -n_0 \right)}; \notag
\end{equation}
as the states in the GQDM Hilbert space are invariant under $G_v$ for all $0 \leq \alpha \leq 2 \pi$, this represents a local $U(1)$ gauge symmetry. The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have taken $t_\bplqt = t$ \& $v_\bplqt =v$ $\forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}

In this case of a fixed number constraint, the Hamiltonian conserves the dimer flux through a closed loop:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0
\end{equation}
and therefore we can define two conserved topological winding numbers $\Phi_{x}$ \& $\Phi_{y}$ which are the dimer flux through topologically non-trival loops that wind around each axis of the lattice when the system has periodic conditions (see \figref{fig:orientedSL}). On an lattice of dimensions $L_x\times L_y$, the the eigenvalues of $\Phi_{x/y}$ integers in range	
\begin{equation}
-L_{x/y}/2 \leq  \Phi_{x/y} \leq L_{x/y}/2,
\end{equation}
thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

@@

The gauge constraint requires each vertex to have $Q_v = 0$; thus defects which violate the constraint have $Q_v \neq 0$. A table of the charges associated with defects in the QDM and QFPLM is listed in Table \ref{tab:defects}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \begin{tabular}{| l | c | c | c |}
  	\hline
      	type & $n_v$ & $Q_{\rm{QDM}}$ & $Q_{\rm{FPLM}}$  \\
	\hline
	monomer & $0$ & -1 & -2 \\
	\hline
	dimer & $1$ & 0 & -1 \\
	\hline
	trimer & $2$ & 1 & 0 \\
	\hline
	tetramer & $3$ & 2 & 1 \\
	\hline
	pentamer & $4$ & 3 & 2 \\
	\hline
  \end{tabular}
  \caption{Defects and their associated charges on the A sub-lattice in both the QDM and QFPLM. Defects on the B sub-lattice have opposite charge.}
  \label{tab:defects}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We now consider relaxing the local constraint to a parity constraint such that $n_v \mod 2 = n_P$ for $n_P = 0,1$; the resulting GQDM are the even \& odd parity Toric Code (or Ising gauge theory). These models have a correspondingly reduced gauge symmetry which can be written as
\begin{equation}
G_v^P = e^{\pm i \pi \left(n_v-n_P\right)}, \notag
\end{equation}
and the gauge symmetry is now reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{TC}}$ is expanded accordingly \textcolor{red}{Owen: Add symbols for 4-0 resonance, and 1-3 resonances and 2-2 resonance}:
\begin{align}
\mathcal{C}_{\rm{TC}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{ \left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right) \bigr\}
\end{align}
where we implicitly include rotational symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HTC\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{TC}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HTC$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

The increased dynamics of $\HTC$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HTC$. Consequently, the string operator which commutes with $\HTC$ (for closed strings) is
\begin{equation}
S_\gamma^{\rm{TC}} \equiv e^{i\pi \Phi_\gamma}.
\end{equation}
Thus for each topologically non-trival loop, $S_\Gamma$ has two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as $\Omega: (\pm 1,\pm1)$. Thus the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

Defects at a vertex in the Toric Code models carry a $Z_2$ electric charge of $-1$, consistent with the $Z_2$ gauge theory description of the models.

    \subsection{$Z_3$ Toric Code on the Square Lattice}
        In the $Z_3$ toric three degrees of freedom live on each link.
        We can represent the $Z_3$ algebra through the following expressions
        %
        %Z_N
        %\begin{equation}
        %    E |n\rangle = n |n\rangle, ~ n=0,1,2,...,N-1
        %    \\
        %    e^{iA} |n\rangle = |n+1 \rangle, ~ n=0,1,2,...,N-2
        %    \\
        %    e^{iA} |N-1\rangle = |0\rangle
        %    %\label{eqn:}
        %\end{equation}
        %Z_3
        %
        \begin{equation}
            \begin{split}
            & E_l |n_l\rangle = n_l |n_l\rangle, ~ n_l=0,1,2
            \\
            & e^{iA_l} |n_l \rangle = |n_l+1 \rangle, ~ n_l =0,1
            \\ 
            & e^{iA_l} |2 \rangle = |0\rangle
            \end{split}
            ,
            %\label{eqn:}
        \end{equation}
        %
        where the operators $E_l$ and $A_l$ act on a given link $l$ and are analogous to the 
        electric field and vector potential respectively. 
        By assigning orientations to the links between two vertices we can define the three degrees of
        freedom pictorially. 
        %
        \begin{itemize}
            \item $\vvemptylink$ 
            \item $\rightarrowlink$ 
            \item $\leftarrowlink$ 
        \end{itemize}
        % 
        To simplify notation, we define $e^{iA_l}$ as the operator $Q$ which acts on the links in the
        following way
        \begin{equation}
            \begin{split}
                & Q^\dagger_{vv'}  | \vvemptylink \rangle = | \rightarrowlink \rangle
                ,
                \\
                & Q^\dagger_{vv'}  | \rightarrowlink \rangle = | \leftarrowlink \rangle
                ,
                \\
                & Q^\dagger_{vv'}  | \leftarrowlink \rangle = | \emptylink \rangle
                .
            \end{split}
            %\label{eqn:}
        \end{equation}
        The link is specified by the two vertices it connects. The operator
        $Q^\dagger_{vv'}$ acts in the direction from the first subscript to the second subscript, in
        this case denoted as $v$ and $v'$. 
        Changing the direction in which $Q$ operates is equivalent to taking the hermitian conjugate
        of the operator withougt changing the direction.  
        %
        \begin{equation}
            (Q^\dagger_{vv'})^\dagger = Q_{vv'}^\dagger
            .
            %\label{eqn:}
        \end{equation}

>>>>

\newcommand{\figref}[1]{Fig. \ref{#1}}
\newcommand{\secref}[1]{Sec. \ref{#1}}
\newcommand{\tabref}[1]{Tab. \ref{#1}}

@@

\newcommand{\HIGT}{H_\mathrm{IGT}}
\newcommand{\HQDPM}{H_\mathrm{QDPM}}

@@

\title{$Z_3$ topological order in the quantum dimer-pentamer model}

@@

Over the last several decades, experimental and theoretical work has demonstrated that strongly interacting quantum many-body systems support exotic quantum phases phases of matter not described by the Landau symmetry breaking paradigm. One class of such quantum phases of matter are topologically ordered quantum liquids~\cite{Wen1990}, first realized experimentally via the fractional quantum Hall effect~\cite{FQHE????}. Despite the absence of conventional symmetry breaking and a local order parameter, topological quantum liquids possess a non-trivial quantum order that distinguishes them from trivial liquid phases~\cite{Nayak2008}. Although not detectable by local operators, topological order arises due to a non-local entanglement structure that leaves a signatures in the bipartite entanglement entropy~\cite{Levin2006a,Kitaev2006b}. Such topological phases may provide the basis for fault-tolerant quantum information processing~\cite{Freedman2001,Kitaev2003}.

Because of the exotic nature of topological phases and their potential use for quantum information processing, there is great interest in the development of theoretical models which display these phases. Such models offer the possibility of furthering the understanding of these phases as well as providing possibly candidate experimental systems in which to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Martinis2015}. A variety of exactly soluble lattice models are known to support ground states with topological order that ranges from the simplest variety, $Z_2$ topological order~\cite{Kitaev2003,Wen2003}, to those with non-Abelian anyonic excitations~\cite{Levin2005a,Kitaev2006a}; such models generally involve complex many-body interactions, and thus it is desirable to find simpler models supporting topological order that are closer to what is accessible experimentally.

Among the simplest class of models displaying topological quantum liquid phases are those with local geometric constraints; the constraints inhibit the formation of symmetry breaking local order and given sufficient dynamics, may lead to a symmetric liquid ground state. One of the most basic of these is the quantum dimer model (QDM) describing dimer degrees of freedom living on the links of a lattice~\cite{Rokhsar1988,Moessner2008}; the hard-core, fully-packed QDM has a local constraint that requires exactly one dimer touching each vertex.

@@

While on non-bipartite lattices, the QDM posses a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the QDM is relaxed to a parity constraint such that each vertex much have an odd (or even) number of dimers touching each vertex, the simplest non-trivial dynamics leads to a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in dimer-based models is intimately related to the local constraint which imposes a local gauge symmetry in the Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local (Ising) Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can often be enforced by a local potential energy penalty and the required dynamics are generally expected to arise at the lowest order non-trivial order in perturbation theory. We will refer to such models with dimer degrees of freedom on links and local constraints at vertices as \emph{generalized quantum dimer modes} (GQDMs). Previous work as demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires only $Q_v \mod 2$ to vanish $\forall v$.

@@

Because of the local constraints, all states $\ket{\psi}$ in the Hilbert space of a GQDM are invariant under local gauge transformations $G_v \ket{\psi} = \ket{\psi}$ a the vertices, $G_v$ of the general form~\cite{Moessner2001}
\begin{equation}
G_v = e^{i \alpha_v Q_v}, \label{eq:Gv}
\end{equation}
where the allowed values of $\alpha_v$ depend on the constraint.  
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

If a defect is introduced which violates the local constraint at a vertex, this can be viewed as an ``electric charge'' living on the vertex, within the context of the original GQDM. 
%Defects may be created in a GQDM by defining an ``electric string'' operator that creates pairs of defects and thus generating ``charged'' states that are outside the gauge invariant Hilbert space:
%\begin{equation}
%S_{vv'}^e \equiv \prod_{\ell \in \tilde{\gamma}_{vv'}} \sigma_\ell^x
%\end{equation}
%where $\tilde{\gamma}_{vv'}$ is an open string on the lattice from $v$ to $v'$. When acting on a gauge invariant state, $S_{\tilde{\gamma}}^e$ creates a neutral pair of charges $Q_v = \pm 1$ at the ends of $\tilde{\gamma}$ where the signs depend on the lattice and the initial configuration of vertices.
One may determine if such electric charges are confined by introducing a pair of such defects and computing their spatial correlation function, thus determining the confined or deconfined nature of the GQDM ground state.

The gauge symmetry allows one to construct a gauge invariant string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv -\frac{c_P}{4} \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) 
\end{equation}
and the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma+ \tilde{\Phi}_\Gamma$. (Alternately can think of the source of $\Phi_\Gamma$ to be static charges of charge $+c_0$ ($-c_0$) sitting at the verticies of the $A$($B$) sub-lattice.).

We define the string operator is defined as
\begin{equation}
S_\gamma \left( \alpha \right) \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)} \label{eq:Sgamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic charges.

On surfaces with non-trivial topology, we may define the topological loop operators
\begin{equation}
W_{X,Y} \left( \alpha \right) \equiv S_{\Gamma_{X,Y}} \left( \alpha \right)
\end{equation}
where $\Gamma_{X,Y}$ are topologically non-trivial closed loops that wind once about the $X$ and $Y$ directions of the torus, respectively. Since $W_{X,Y} (\alpha)$ commute with the Hamiltonian, we may use their eigenvalues to label the ground states.

@@

The two unique fixed-number constraints on the square lattice are lead to QDM in the odd sector and the quantum full-packed loop model (QFPLM) in the even sector. For configurations that satisfy these a number constraints, $Q_v$ strictly vanishes, and thus all states in the Hilbert space are invariant under $G_v$ for $0 \leq \alpha_v \leq 2 \pi$ in \Eqref{eq:Gv}; this represents a local $U(1)$ gauge symmetry.

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have taken $t_\bplqt = t$ \& $v_\bplqt =v$ $\forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0
\end{equation}
and therefore we can define two conserved topological winding numbers $\Phi_{x}$ \& $\Phi_{y}$ which are the dimer flux through topologically non-trival loops that wind around each axis of the lattice when the system has periodic conditions (see \figref{fig:orientedSL}). On an lattice of dimensions $L_x\times L_y$, the the eigenvalues of $\Phi_{x/y}$ integers in range	
\begin{equation}
-L_{x/y}/2 \leq  \Phi_{x/y} \leq L_{x/y}/2,
\end{equation}
thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

@@

Defects in these models correspond to states with $Q_v \neq 0$; a table of the charges associated with defects in the QDM and QFPLM is listed in \tabref{tab:defects}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \begin{tabular}{| l | c | c | c |}
  	\hline
      	type & $n_v$ & $Q_v^{\rm{QDM}}$ & $Q_v^{\rm{FPLM}}$  \\
	\hline
	monomer & $0$ & -1 & -2 \\
	\hline
	dimer & $1$ & 0 & -1 \\
	\hline
	trimer & $2$ & 1 & 0 \\
	\hline
	tetramer & $3$ & 2 & 1 \\
	\hline
	pentamer & $4$ & 3 & 2 \\
	\hline
  \end{tabular}
  \caption{Defects where $n_v$ dimers touch a vertex and their associated $U(1)$ charges on the A sub-lattice in both the QDM and QFPLM. Defects on the B sub-lattice have opposite charge.}
  \label{tab:defects}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) (and the closely related even and odd sector Toric Codes). In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{TC}}$ is expanded accordingly \textcolor{red}{Owen: Add symbols for 4-0 resonance, and 1-3 resonances and 2-2 resonance}:
\begin{align}
\mathcal{C}_{\rm{TC}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{ \left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right) \bigr\}
\end{align}
where we implicitly include rotational symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

The increased dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators which commutes with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{\rm{IGT}} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{\rm{IGT}} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{\rm{IGT}}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as $\Omega: (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

Defects in the Ising Gauge theories have $Q_v^{Z_2} = 1$ and we identify these defects as $Z_2$ electric charges on the corresponding vertices.

The ground states of $\HIGT$ are fourfold degenerate on the torus and have exponentially decaying dimer-dimer and monomer-monomer correlations (\textcolor{red}{Is this known for both sectors}) and a finite energy gap; such features demonstrated the deconfined topologically ordered nature of the ground state of the IGTs.

We may understand the emergence of the $Z_2$ topologically ordered phase of IGTs from the QDM and QFPLM from the following picture. The additional vertex configurations that are allowed in the IGTs (monomers and pentamers in the even sector and tetramers in the odd sector) relative to the number constrained models have $Q_v = \pm 2$. It is known that coupling a $U(1)$ gauge field to a charge $q$ matter field can result in an $Z_q$ gauge theory~\cite{Fradkin1979}; thus in the IGTs we have addded dynamic charge 2 defects to the underlying $U(1)$ symmetric model, and we may expect a $Z_2$ gauge theory to result. We note that pentamers are charge $\pm3$ objects in the QDM (see \tabref{tab:defects}); thus below we investigate whether introducing pentamers into the QDM leads to an effective $Z_3$ gauge theory.

We define the Hilbert space of the QDPM to be an odd sector GQDM where vertices are constrained to be touched by a single dimer or the center of a pentamer, such that $n_v = 1,4$. For defect-free configurations in the QDPM, $Q_v$ only vanishes modulo $3$;  thus we may define a $Z_3$ charge,
\begin{equation}
Q_v^{Z_3} \equiv \left(Q_v~\rm{mod}~3\right)
\end{equation}
 which strictly vanishes at each vertex, representing the local constraint. All states in the QDPM are invariant under an exact local $Z_3$ gauge symmetry given by \Eqref{eq:Gv} with $\alpha_v \in \{0,\pm 2\pi/3\}$. Thus introducing pentamers in the QDM reduces the gauge symmetry from $U(1)$ to $Z_3$.

The Hamiltonian of the QDPM is defined by \Eqref{eq:HGQDM} including both plaquette flips as well as local dimer resonances that create and destroy pairs of pentamers, and allow single pentamers to translate on the same sub-lattice (see \figref{fig:pentamer_moves}). Thus we may write $\HQDPM$ as the following:
\begin{equation}
\HQDPM \equiv \HQDM + H_{\rm{pent}}
\end{equation}
where $H_{\rm{pent}}$ only involves the resonances in \figref{fig:pentamter_moves} involving pentamers.

 %%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
\centering
\includegraphics[width=0.8\linewidth]{pentamer_moves.pdf}
\caption{a) Horizontal creation an annihilation, b) horizontal move \textcolor{red}{Owen:fix this move}, c) diagonal moves \textcolor{red}{Owen: Remove (d) as it is related to (c) by rotational symmetry}}
\label{fig:pentamer_moves}
\end{figure}
%%%%%%%%%%%%%%%%%

Because of the terms involving pentamers, $\HQDPM$ conserves $Q_v^{Z_3}$ rather than $Q_v$, and thus $\Phi_\Gamma$ does not commute with the Hamiltonian for closed loops $\Gamma$. The appropriate string and loop operator corresponds to \Eqref{eq:Sgamma} with $\alpha = 2\pi/3$:
\begin{equation}
S_\gamma^{\rm{QDPM}} \equiv S_\gamma \left( 2\pi/3\right), \quad W_{X,Y}^{\rm{QDPM}} \equiv W_{X,Y} \left( 2\pi/3\right).
\end{equation}
Thus we have 
\begin{equation}
\left[ W_{X,Y}^{\rm{QDPM}}, \HQDPM \right] = 0.
\end{equation}
and for closed curves $\Gamma$
\begin{equation}
\left[ S_\Gamma^{\rm{QDPM}}, \HQDPM \right] = 0.
\end{equation}
